---
title: "Tumor-only variant analysis report"
params:
  run_dir: "."
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
run_dir <- params$run_dir
`%||%` <- function(a, b) if (is.null(a) || length(a) == 0) b else a
rd <- function(name) { p <- file.path(run_dir, "tables", name)
  if (file.exists(p)) as.data.frame(data.table::fread(p, sep = "\t", na.strings = c("", ".", "NA"))) else NULL }
manifest <- tryCatch(jsonlite::read_json(file.path(run_dir, "manifest.json"), simplifyVector = TRUE),
                     error = function(e) list())
v_all <- rd("variants_all.tsv.gz"); audit <- rd("filter_audit.tsv.gz")
retained <- rd("variants_retained.tsv.gz"); excluded <- rd("variants_excluded.tsv.gz")
top <- rd("top_prioritized_variants.tsv"); inv <- rd("predictor_inventory.tsv")
conf <- rd("predictor_conflicts.tsv"); ctx <- rd("cosmic_context_matches.tsv")
schema <- rd("input_schema_report.tsv"); oncokb <- rd("oncokb_annotations.tsv.gz")
plotman <- tryCatch(as.data.frame(data.table::fread(file.path(run_dir, "plots", "plot_manifest.tsv"))),
                    error = function(e) NULL)
has_dt <- requireNamespace("DT", quietly = TRUE)
resolve_plot_path <- function(path) {
  if (length(path) == 0 || is.na(path) || !nzchar(path)) return(path)
  is_abs <- grepl("^(/|[A-Za-z]:[/\\\\])", path)
  if (is_abs && file.exists(path)) return(normalizePath(path, mustWork = FALSE))
  if (file.exists(path)) return(normalizePath(path, mustWork = FALSE))
  if (grepl("/plots/", path, fixed = TRUE)) {
    rel <- sub("^.*?/plots/", "", path)
    candidate <- file.path(run_dir, "plots", rel)
    if (file.exists(candidate)) return(normalizePath(candidate, mustWork = FALSE))
  }
  candidate <- file.path(run_dir, "plots", basename(path))
  if (file.exists(candidate)) return(normalizePath(candidate, mustWork = FALSE))
  path
}
g <- function(x, d = "NA") if (is.null(x) || length(x) == 0 || is.na(x[1])) d else x[1]
cnt <- function(cond) if (is.null(v_all) || !length(cond)) 0 else sum(cond, na.rm = TRUE)
n_input <- g(manifest$n_input_variants, if (!is.null(v_all)) nrow(v_all) else 0)
n_ret <- if (!is.null(retained)) nrow(retained) else 0
n_exc <- if (!is.null(excluded)) nrow(excluded) else 0
n_rev <- cnt(v_all$filter_status == "REVIEW")
n_high <- if (!is.null(v_all) && "PRIORITY_CATEGORY" %in% names(v_all)) cnt(v_all$PRIORITY_CATEGORY == "high") else 0
n_hot <- if (!is.null(v_all) && "hotspot_match" %in% names(v_all)) cnt(v_all$hotspot_match %in% c(TRUE, "TRUE")) else 0
n_comp <- if (!is.null(v_all) && "COMPUTATIONAL_EVIDENCE_CATEGORY" %in% names(v_all)) cnt(v_all$COMPUTATIONAL_EVIDENCE_CATEGORY %in% c("strong_support","moderate_support")) else 0
n_conf <- if (!is.null(conf)) nrow(conf) else 0
n_ctx_same <- if (!is.null(ctx) && "COSMIC_TUMOR_CONTEXT_STATUS" %in% names(ctx)) sum(ctx$COSMIC_TUMOR_CONTEXT_STATUS %in% c("exact_match","compatible_match","pan_cancer_with_sample_tumor"), na.rm = TRUE) else 0
n_artifact <- if (!is.null(v_all) && "VARIANT_AUTHENTICITY_CATEGORY" %in% names(v_all)) cnt(v_all$VARIANT_AUTHENTICITY_CATEGORY == "likely_artifact") else 0
n_true <- if (!is.null(v_all) && "VARIANT_AUTHENTICITY_CATEGORY" %in% names(v_all)) cnt(v_all$VARIANT_AUTHENTICITY_CATEGORY == "likely_true") else 0
card <- function(label, value, color = "#2c7fb8") sprintf(
  "<div style='display:inline-block;min-width:150px;margin:6px;padding:12px 16px;border-radius:10px;background:%s;color:#fff;text-align:center'><div style='font-size:26px;font-weight:700'>%s</div><div style='font-size:12px;opacity:.9'>%s</div></div>",
  color, value, label)
```

# Executive summary {-}

```{r, results='asis'}
cat("<div>",
  card("Input variants", n_input, "#34495e"),
  card("Retained", n_ret, "#27ae60"),
  card("Excluded", n_exc, "#c0392b"),
  card("Manual review", n_rev, "#e67e22"),
  card("High priority", n_high, "#8e44ad"),
  card("In hotspots", n_hot, "#16a085"),
  card("Computational support", n_comp, "#2980b9"),
  card("COSMIC same-tumor", n_ctx_same, "#d35400"),
  card("Likely true", n_true, "#1e8449"),
  card("Likely artifact", n_artifact, "#7f8c8d"),
  "</div>", sep = "\n")
```

**Run:** `r g(manifest$run_id)` &nbsp;|&nbsp; **Build:** `r g(manifest$genome_build)` &nbsp;|&nbsp;
**Input format:** `r g(manifest$input_format)` &nbsp;|&nbsp; **COSMIC:** `r g(manifest$cosmic_release)` &nbsp;|&nbsp;
**OncoKB:** `r if (is.null(oncokb)) "not run" else "annotation only (does not affect filtering)"` &nbsp;|&nbsp;
**Package:** `r g(manifest$package_version)`

## Principais achados

```{r, results='asis'}
findings <- c(
  sprintf("- **%d** variant(s) classified as **high priority**.", n_high),
  sprintf("- **%d** variant(s) in known **hotspots**.", n_hot),
  sprintf("- **%d** variant(s) with concordant **computational support** from multiple predictors.", n_comp),
  sprintf("- **%d** variant(s) with **conflicting** predictor evidence (e.g. SIFT vs PolyPhen).", n_conf),
  sprintf("- **%d** variant(s) with COSMIC recurrence in the **same tumor context**.", n_ctx_same),
  sprintf("- **%d** variant(s) assessed as **likely true** and **%d** as **likely artifact** (authenticity score).", n_true, n_artifact),
  sprintf("- **%d** variant(s) require **manual review** (uncertain / possible germline).", n_rev))
cat(paste(findings, collapse = "\n"))
```

## Alerts

```{r, results='asis'}
alerts <- character()
add_alert <- function(cond, msg) if (isTRUE(cond)) alerts[[length(alerts) + 1]] <<- msg
add_alert(is.na(g(manifest$genome_build, NA)), "Genome build could not be confirmed.")
add_alert(n_ret == 0, "No variants were retained.")
add_alert(is.null(manifest$cosmic_release) || is.na(g(manifest$cosmic_release, NA)), "No COSMIC database configured — COSMIC evidence unavailable.")
if (!is.null(schema)) { miss <- schema[schema$required & !schema$present, "canonical_field"]
  add_alert(length(miss) > 0, paste("Missing required canonical field(s):", paste(miss, collapse = ", "))) }
if (!is.null(inv)) { det <- sum(inv$status == "detected", na.rm = TRUE)
  add_alert(det == 0, "No functional predictors were detected in the input.") }
if (!is.null(v_all) && "COSMIC_CONTEXT_EVALUABLE" %in% names(v_all))
  add_alert(any(v_all$COSMIC_MATCH %in% c(TRUE,"TRUE")) && all(!(v_all$COSMIC_CONTEXT_EVALUABLE %in% c(TRUE,"TRUE"))),
            "COSMIC matches present but tumor context is not evaluable (no/unmapped tumor type).")
if (!is.null(plotman)) { sk <- plotman[plotman$status == "skipped", ]
  if (nrow(sk)) for (i in seq_len(nrow(sk))) add_alert(TRUE, sprintf("Plot '%s' not generated: %s", sk$plot[i], sk$reason[i])) }
if (length(alerts) == 0) cat("_No critical alerts._\n") else
  cat(paste(sprintf("- ⚠️ %s", alerts), collapse = "\n"))
```

# Top prioritized variants

*What this shows:* the highest-priority retained variants for review. *How to
interpret:* priority is a decomposed heuristic score (consequence + computational
+ hotspot + driver + clinical + COSMIC global); it ranks, it does not filter.

```{r}
if (!is.null(top) && nrow(top) > 0) {
  show <- utils::head(top, 200)
  if (has_dt) DT::datatable(show, filter = "top", rownames = FALSE,
    extensions = "Buttons", options = list(pageLength = 15, scrollX = TRUE,
      dom = "Bfrtip", buttons = c("copy", "csv")), class = "compact stripe")
  else knitr::kable(utils::head(show, 50))
} else "No retained variants to prioritize."
```

# Details {.tabset}

## Data quality

```{r}
if (!is.null(schema)) { if (has_dt) DT::datatable(schema, rownames = FALSE, options = list(pageLength = 10, scrollX = TRUE)) else knitr::kable(schema) }
```

## Filtering funnel

*What this shows:* how many variants pass each stage. Nothing is deleted — excluded
variants remain in `variants_all` with their full decision trail.

```{r}
if (!is.null(audit)) knitr::kable(audit)
```

## Functional predictors

*What this shows:* which in-silico predictors were detected and their coverage.
*How to interpret:* predictors estimate possible functional impact but must not be
read in isolation as evidence of pathogenicity; they never remove a variant.

```{r}
if (!is.null(inv)) { if (has_dt) DT::datatable(inv, rownames = FALSE, options = list(pageLength = 10, scrollX = TRUE)) else knitr::kable(inv) }
```

```{r, results='asis'}
if (!is.null(conf) && nrow(conf) > 0) cat(sprintf("\n**%d variant(s) with conflicting predictors** (see tables/predictor_conflicts.tsv).\n", nrow(conf)))
```

## COSMIC evidence

*What this shows:* global COSMIC recurrence (across all tumor types) vs contextual
recurrence (same/compatible tumor type). *How to interpret:* global recurrence is
informational; only tumor-context matches inform OS4/OM4. The contextual score is
**experimental** — the main output uses `CONFIDENCE_SCORE_BASE`.

```{r}
if (!is.null(ctx) && nrow(ctx) > 0) { if (has_dt) DT::datatable(utils::head(ctx, 200), rownames = FALSE, options = list(pageLength = 10, scrollX = TRUE)) else knitr::kable(utils::head(ctx, 50)) } else "No COSMIC matches."
```

## maftools visualizations

```{r, results='asis'}
if (!is.null(plotman)) {
  ok <- plotman[plotman$status == "ok", ]
  for (i in seq_len(nrow(ok))) {
    fs <- strsplit(ok$files[i], ";", fixed = TRUE)[[1]]
    png <- fs[grepl("[.]png$", fs)]
    if (length(png)) cat(sprintf("\n### %s\n\n![](%s)\n\n", ok$plot[i], resolve_plot_path(png[1])))
    else cat(sprintf("\n### %s — %s\n", ok$plot[i], paste(basename(fs), collapse = ", ")))
  }
} else cat("No maftools plots produced.\n")
```

## Excluded variants

```{r}
if (!is.null(excluded) && nrow(excluded) > 0) {
  cols <- intersect(c("sample_id","gene","consequence","final_class","filter_status","filters_failed","filter_reasons"), names(excluded))
  ex <- excluded[, cols, drop = FALSE]
  if (has_dt) DT::datatable(utils::head(ex, 500), filter = "top", rownames = FALSE, options = list(pageLength = 10, scrollX = TRUE)) else knitr::kable(utils::head(ex, 50))
} else "No excluded variants."
```

## Provenance & parameters

```{r}
knitr::kable(data.frame(field = names(unlist(manifest)), value = as.character(unlist(manifest))))
```

# Limitations {-}

- Tumor-only analysis **cannot definitively distinguish somatic from germline** variants.
- Functional predictors are **supporting evidence only** and never remove a variant.
- COSMIC **global** recurrence is informational; **contextual** COSMIC evidence and the
  `COSMIC_CONTEXT_SUPPORT_SCORE` are **experimental**.
- OS4/OM4 tumor-context gating and priority weights are **heuristic**, not clinically validated.
- OncoKB (if run) is a confirmatory annotation and never changes filtering.
- Results require review by qualified professionals. This is **not a validated clinical test**.
```
