---
title: "Tumor-only variant analysis report"
params:
  run_dir: "."
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
run_dir <- params$run_dir
read_tbl <- function(name) {
  p <- file.path(run_dir, "tables", name)
  if (file.exists(p)) data.table::fread(p, sep = "\t", data.table = FALSE, na.strings = c("", ".", "NA")) else NULL
}
manifest <- tryCatch(jsonlite::read_json(file.path(run_dir, "manifest.json"), simplifyVector = TRUE),
                     error = function(e) list())
oncokb <- read_tbl("oncokb_annotations.tsv.gz")
variants_all <- read_tbl("variants_all.tsv.gz")
retained <- read_tbl("variants_retained.tsv.gz")
excluded <- read_tbl("variants_excluded.tsv.gz")
audit <- read_tbl("filter_audit.tsv.gz")
cosmic <- read_tbl("cosmic_matches.tsv.gz")
g <- function(x, d = "NA") if (is.null(x) || length(x) == 0) d else x
```

## 1. Execution summary

- **Run id:** `r g(manifest$run_id)`
- **Date:** `r g(manifest$date)`
- **Package version:** `r g(manifest$package_version)`
- **Genome build:** `r g(manifest$genome_build)`
- **Reference FASTA:** `r g(manifest$fasta_file)`
- **COSMIC release:** `r g(manifest$cosmic_release)`
- **Input:** `r g(manifest$input_file)` (`r g(manifest$n_input_variants)` variants)
- **OncoKB annotation:** `r if (is.null(oncokb)) "not run" else "present (annotation only — does not affect filtering)"`
- **Runtime (s):** `r g(manifest$runtime_seconds)`

## 2. Provenance

- **Input SHA-256:** `r g(manifest$input_sha256)`
- **R version:** `r g(manifest$r_version)`
- **Key packages:** `r paste(names(unlist(manifest$packages)), unlist(manifest$packages), sep = "=", collapse = ", ")`

```{r}
utils::sessionInfo()$R.version$version.string
```

## 3. Input QC

```{r}
if (!is.null(variants_all)) {
  num <- function(col) if (col %in% names(variants_all)) suppressWarnings(as.numeric(variants_all[[col]])) else NA_real_
  data.frame(
    metric = c("n variants", "median DP", "median VAF", "% missing gene",
               "% missing consequence", "n distinct samples"),
    value = c(nrow(variants_all),
              round(stats::median(num("dp"), na.rm = TRUE), 1),
              round(stats::median(num("vaf"), na.rm = TRUE), 3),
              round(100 * mean(is.na(variants_all$gene) | variants_all$gene == ""), 1),
              round(100 * mean(is.na(variants_all$consequence) | variants_all$consequence == ""), 1),
              length(unique(variants_all$sample_id))))
}
```

## 4. Filtering funnel

Every variant is retained in `variants_all` with its full decision trail; the
counts below are views, not deletions.

```{r}
audit
```

## 5. Retained variants

```{r}
if (!is.null(retained) && nrow(retained) > 0) {
  cols <- intersect(c("sample_id","gene","protein_change","consequence","vaf","dp",
    "max_pop_af","COSMIC_EVIDENCE_SUMMARY","confidence_category","confidence_score",
    "final_class","driver_class","filter_reasons"), names(retained))
  retained[, cols, drop = FALSE]
} else "No retained variants."
```

## 6. Excluded variants

```{r}
if (!is.null(excluded) && nrow(excluded) > 0) {
  cols <- intersect(c("sample_id","gene","protein_change","consequence",
    "final_class","filter_status","filters_failed","filter_reasons"), names(excluded))
  excluded[, cols, drop = FALSE]
} else "No excluded variants."
```

## 7. Visualizations

```{r, results='asis'}
onco <- file.path(run_dir, "plots", "oncoplot_top_genes.pdf")
if (file.exists(onco)) cat(sprintf("Oncoplot (maftools): [%s](%s)\n\n", basename(onco), onco)) else
  cat("Oncoplot not produced (insufficient retained variants or maftools unavailable).\n\n")
```

## 7a. Evidência COSMIC global

Global COSMIC recurrence across ALL tumor types (a transformed raw recurrence
count, not a frequency). Global evidence is informative but is **not** somatic
support in the analyzed tumor and does **not** grant OS4/OM4.

```{r}
if (!is.null(cosmic) && nrow(cosmic) > 0 && "COSMIC_TOTAL_OCCURRENCES" %in% names(cosmic)) {
  gcols <- intersect(c("sample_id","gene","COSMIC_TOTAL_OCCURRENCES",
    "COSMIC_GLOBAL_RECURRENCE_SCORE","COSMIC_GLOBAL_RECURRENT",
    "COSMIC_GLOBAL_PANCANCER_RECURRENT","COSMIC_GLOBAL_EVIDENCE_ONLY"), names(cosmic))
  head(cosmic[, gcols, drop = FALSE], 300)
} else "No COSMIC matches."
```

## 7b. Evidência COSMIC no contexto do tipo tumoral

COSMIC evidence stratified by the sample's tumor type. Occurrences in the sample's
(harmonized) tumor type are separated from those in other tumor types; raw counts
are interpreted with a base-size-aware transform and are **not** compared directly
across tumor types. OS4/OM4 are granted only from tumor-context occurrences. See
`docs/COSMIC_DECISION_RULES.md` and `docs/COSMIC_OS4_OM4_AUDIT.md`.

```{r}
if (!is.null(cosmic) && nrow(cosmic) > 0) {
  by_status <- as.data.frame(table(cosmic$COSMIC_TUMOR_CONTEXT_STATUS))
  names(by_status) <- c("tumor_context_status", "n_variants")
  by_status
} else "No COSMIC matches (no processed COSMIC database configured, or no genomic matches)."
```

```{r}
if (!is.null(cosmic) && nrow(cosmic) > 0) {
  cols <- intersect(c("sample_id","gene","tumor_type","tumor_subtype",
    "COSMIC_TUMOR_CONTEXT_STATUS","COSMIC_MATCH_LEVEL","COSMIC_CONTEXT_EVALUABLE",
    "COSMIC_CONTEXT_INTERPRETATION","COSMIC_MATCHING_TUMOR_OCCURRENCES",
    "COSMIC_OTHER_TUMOR_OCCURRENCES","COSMIC_MATCHED_TUMOR_RECURRENCE_SCORE",
    "COSMIC_TUMOR_SPECIFICITY_SCORE","COSMIC_CONTEXT_SUPPORT_SCORE",
    "OS4_CONTEXTUAL","OM4_CONTEXTUAL","COSMIC_GLOBAL_EVIDENCE_ONLY",
    "COSMIC_PANCANCER_INCLUDES_SAMPLE_TUMOR","COSMIC_TUMOR_CONTEXT_REASON"), names(cosmic))
  head(cosmic[, cols, drop = FALSE], 300)
}
```

The COSMIC context score (`COSMIC_CONTEXT_SUPPORT_SCORE`) is **experimental /
complementary**. Main filtering and confidence use `CONFIDENCE_SCORE_BASE`, which
does not include tumor-context COSMIC evidence.

```{r, results='asis'}
if (!is.null(cosmic) && nrow(cosmic) > 0) {
  unk <- sum(cosmic$COSMIC_TUMOR_CONTEXT_STATUS == "tumor_type_unknown", na.rm = TRUE)
  miss <- sum(cosmic$COSMIC_TUMOR_CONTEXT_STATUS == "cosmic_tumor_type_missing", na.rm = TRUE)
  if (unk > 0) cat(sprintf("- **Alert:** %d matched variant(s) from samples whose tumor type could not be harmonized.\n", unk))
  if (miss > 0) cat(sprintf("- **Alert:** %d COSMIC record(s) lacked tumor-context information.\n", miss))
  cat("- **Bias note:** COSMIC representation is uneven across tumor types; absence in a tumor type does not exclude relevance, and cross-tumor counts are not directly comparable.\n")
}
```

## 8. OncoKB annotation (confirmatory only)

```{r}
if (!is.null(oncokb)) {
  cols <- intersect(c("sample_id","gene","ONCOKB_ONCOGENIC","ONCOKB_MUTATION_EFFECT",
    "ONCOKB_HIGHEST_LEVEL","ONCOKB_QUERY_STATUS"), names(oncokb))
  head(oncokb[, cols, drop = FALSE], 200)
} else "OncoKB annotation was not run. It is optional and never affects filtering."
```

## 9. Limitations

- Tumor-only analysis **cannot definitively distinguish somatic from germline** variants.
- **Presence in COSMIC is supporting evidence** of recurrence, not proof of somaticity.
- **Absence from COSMIC does not exclude** clinical or biological relevance.
- **OncoKB was used only as a confirmatory annotation** and never changed any filtering decision.
- Results must be reviewed by qualified professionals. This pipeline is **not a validated clinical test**.
