---
title: "tumoronly v2 reproducible tutorial"
author: "Mariana Boroni lab"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{tumoronly v2 reproducible tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Scope

This vignette mirrors `TUTORIAL.md`. It documents the current v2 workflow:

```text
init -> validate -> run -> report
```

Tumor-only classes are probabilistic candidates. They should not be reported as
confirmed clinical somatic or germline calls without matched-normal or
orthogonal validation.

# Source-checkout run

From the repository root:

```{bash validate-demo, eval=FALSE}
Rscript exec/tumoronly validate \
  --input data/demo/variants.tsv \
  --config config/demo_v2.yml \
  --output results/demo_v2_validation.html
```

```{bash run-demo, eval=FALSE}
Rscript exec/tumoronly run \
  --input data/demo/variants.tsv \
  --config config/demo_v2.yml \
  --output results/demo_v2 \
  --run-id tutorial_demo
```

```{bash report-demo, eval=FALSE}
Rscript exec/tumoronly report \
  --run-dir results/demo_v2/tutorial_demo \
  --output results/demo_v2/tutorial_demo/report.html
```

# Installed-package run

```{r installed-demo, eval=FALSE}
library(tumoronly)
library(yaml)

variants <- system.file("extdata", "demo_variants.tsv", package = "tumoronly")
metadata <- system.file("extdata", "demo_sample_metadata.tsv", package = "tumoronly")

cfg <- tumoronly_default_config()
cfg$input$path <- variants
cfg$input$genome_build <- "GRCh38"
cfg$input$sample_metadata <- metadata
cfg$analysis$output_dir <- "results/installed_demo"
yaml::write_yaml(cfg, "installed_demo_config.yml")

validation <- validate_tumoronly_input(
  config = "installed_demo_config.yml",
  output = "installed_demo_validation.html"
)
stopifnot(isTRUE(validation$ok))

result <- run_tumoronly(
  config = "installed_demo_config.yml",
  run_id = "installed_demo"
)

result$run_dir
```

# Outputs

Each v2 run writes:

```text
report.html
run_manifest.json
config_used.yaml
session_info.txt
logs/warnings.tsv
tables/classified_variants.tsv
tables/sample_summary.tsv
tables/filter_audit.tsv
tables/high_confidence_somatic.tsv
tables/likely_germline.tsv
tables/likely_artifact.tsv
tables/known_drivers.tsv
figures/*.pdf
figures/*.svg
figures/*.png
figure_data/*.tsv
```

The key per-variant evidence columns are:

- `evidence_supporting_classification`;
- `evidence_against_classification`;
- `missing_evidence`;
- `classification_explanation`.

# Real-data gate

The demo is not the validation gate. The v2 workflow was run on
`data/test/WGS_all_patients_first100k.tsv`; details are recorded in
`docs/REAL_DATA_VALIDATION.md`.

