---
title: "Using the tumor-only filtering pipeline from VEP or ANNOVAR annotated VCFs"
author:
  - name: "Mariana Boroni lab"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Using VEP or ANNOVAR annotated VCFs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

project_root <- normalizePath(getwd(), mustWork = FALSE)
if (!dir.exists(file.path(project_root, "scripts"))) {
  candidate_root <- normalizePath(file.path(getwd(), ".."), mustWork = FALSE)
  if (dir.exists(file.path(candidate_root, "scripts"))) {
    project_root <- candidate_root
  }
}
knitr::opts_knit$set(root.dir = project_root)
```

# Goal

This tutorial shows how to run the pipeline starting from an annotated VCF
generated by VEP, SnpEff, or ANNOVAR. The pipeline accepts either:

- a tabular TSV/CSV file already exported from VEP/ANNOVAR; or
- a `.vcf` / `.vcf.gz` file with annotation fields in `INFO`.

The output is a tumor-only prioritization package with:

- ranked somatic candidates;
- likely germline/polymorphism/artifact removals;
- manual-review candidates;
- oncoplots and QC figures;
- an HTML dashboard.

# 1. Prepare the annotated VCF

## VEP VCF

For VEP, keep the `CSQ` field in the VCF. The pipeline reads the `CSQ` header
and extracts common fields such as `SYMBOL`, `Consequence`, `HGVSp`,
`Existing_variation`, `CLIN_SIG` and population AF fields when present.

Recommended VEP annotation concepts:

| Concept | Useful VEP field |
|---|---|
| Gene | `SYMBOL` |
| Consequence | `Consequence` |
| Protein change | `HGVSp` |
| Known variant | `Existing_variation` |
| ClinVar | `CLIN_SIG` |
| Population AF | `gnomADe_AF`, `gnomADg_AF`, `MAX_AF` |

## ANNOVAR VCF

For ANNOVAR, the pipeline looks for common `INFO` keys such as:

| Concept | Common ANNOVAR key |
|---|---|
| Gene | `Gene.refGene` |
| Consequence | `ExonicFunc.refGene`, `Func.refGene` |
| Protein change | `AAChange.refGene` |
| Population AF | `gnomAD_AF`, `AF_popmax`, `GMAF` |

# 2. Create a config file

Copy an existing YAML config and point `input.variants` to the annotated VCF:

```yaml
input:
  variants: "data/my_tumor_only_vep.vcf.gz"
  delimiter: "\t"
  sample_metadata: "data/sample_metadata.tsv"
  metadata_delimiter: "\t"

output:
  dir: "results/my_project_filter"

cancer:
  default_tumor_type: "UCEC"

hard_filters:
  assay: "WGS"

population_filters:
  use_population_columns:
    - gnomADe_AF
    - gnomADg_AF
    - MAX_AF
    - ABraOM_AF
    - BIPMed_AF
```

When the input ends in `.vcf` or `.vcf.gz`, the pipeline automatically uses the
VCF parser. For tabular VEP/ANNOVAR exports, use the correct delimiter and keep
the original annotation columns.

# 3. Add sample metadata

Sample metadata is strongly recommended:

```text
sample_id	tumor_type	assay	purity
Tumor_001	UCEC	WGS	0.60
Tumor_002	BRCA	WES	0.45
```

At minimum, provide:

| Column | Why it matters |
|---|---|
| `sample_id` | Must match the VCF sample name or tabular sample column. |
| `tumor_type` | Enables tumor-type recurrence and tumor-specific ML. |
| `assay` | Helps interpret depth, target territory and TMB. |
| `purity` | Improves clonality estimates when available. |

# 4. Run the pipeline

```{bash run-example, eval=FALSE}
Rscript scripts/run_automated_variant_analysis.R --config config/my_project.yml
```

The same command works for VEP VCF, ANNOVAR VCF, or TSV/CSV exports.

# 5. What the pipeline does

| Step | What happens |
|---|---|
| Read input | VCF `INFO`/`FORMAT` or tabular columns are standardized. |
| Technical features | Depth, alt count, VAF, TLOD, mapping/base quality and filters are extracted. |
| Population features | Maximum AF is computed from configured global and Brazilian columns. |
| Hard filters | Low support, low depth and technical failures are flagged. |
| Recurrence | Internal recurrence is used as a pseudo-panel of normals. |
| Validation | OncoKB/COSMIC/hotspot references add supporting evidence. |
| Guideline evidence | ClinGen/CGC/VICC-inspired somatic and ACMG/AMP-inspired germline codes are computed. |
| ML | Incremental pan-cancer and tumor-specific models refine prioritization when labels exist. |
| Ranking | Variants are ordered as somatic candidates, likely removals or manual review. |
| TMB/clonality/ancestry | Per-sample summaries and figures are generated. |
| Dashboard | An HTML dashboard summarizes the run. |

# 6. Main outputs

| File | Use |
|---|---|
| `variant_filtering_dashboard.html` | First file to open after a run. |
| `mutation_confidence_ranking.tsv` | Ranked list with somatic, germline, polymorphism and artifact confidence scores. |
| `variants_final.tsv` | Complete per-variant table. |
| `variants_high_confidence_somatic.tsv` | Strict somatic candidate list. |
| `variants_uncertain_for_review.tsv` | Manual-review queue. |
| `removed_likely_germline.tsv` | Variants removed from somatic list as likely germline/polymorphism. |
| `removed_likely_artifact.tsv` | Variants removed from somatic list as likely artifact/technical failure. |
| `figures/oncoplot_confidence_buckets.png` | Oncoplot colored by confidence bucket. |
| `figures/oncoplot_fallback_top_genes.png` | Somatic candidate oncoplot. |

# 7. Interpreting the ranking

`mutation_confidence_ranking.tsv` contains:

| Column | Meaning |
|---|---|
| `somatic_confidence_score` | Support for keeping as a somatic candidate. |
| `germline_confidence_score` | Support for germline origin or germline-like behavior. |
| `polymorphism_confidence_score` | Support for common/benign population polymorphism. |
| `artifact_confidence_score` | Support for technical artifact. |
| `manual_review_priority_score` | Conflict or high-value uncertainty. |
| `recommended_variant_action` | `keep_somatic_candidate`, `remove_from_somatic_list`, or `manual_review`. |
| `confidence_bucket` | More detailed reason-oriented bucket. |

Practical interpretation:

| Action | Interpretation |
|---|---|
| `keep_somatic_candidate` | Keep for downstream somatic analyses, with tumor-only caveat. |
| `remove_from_somatic_list` | Remove from somatic list; usually polymorphism or artifact. |
| `manual_review` | Check IGV, population AF, annotation, recurrence and possible germline implications. |

# 8. Recommended review loop

1. Open `variant_filtering_dashboard.html`.
2. Review the top `manual_review` variants.
3. Add labels to `review_labels.tsv`.
4. Rerun the pipeline.
5. Track whether ML confidence and manual-review burden improve.

# 9. Session info

```{r session-info}
sessionInfo()
```
